---
title: "Geogrphical Coordinates Recorded in 2006 and 2014 in Mwanza Gulf, Lake Victoria"
author: ""
date: ""
output: 
    bookdown::html_document2
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Introduction
This document aimed to help us understand clearly the locations of geographical points marked in 2006 and 2014 across the Mwanza Gulf [@kishe]. Several locations were recorded with a GPS from Butimba Bay to Kissenda Island---making a transect that cross the mouth of the Mwanza Bay. Because the data were recorded at different epoches, we look at each data sources and check whether the data real present the position on grounds 

```{r, quit = TRUE, warning=FALSE, message=FALSE, comment=""}

require(tidyverse)
require(sf) 
require(leaflet)
require(mapview)
require(parzer)
require(sp)
```

# Data
## Sampled Stations 2014
Table \@ref(tab:tab1) shows the original data from the 2014 dataset. Lookin on the variable you notice that variable names of longitude and latitude are swapped! The longitude variable are actually the latitude and viceversa. 

```{r}
gulf = readxl::read_excel("Coordinates Mwanza Gulf_2006, 2008, 2014-2.xlsx", sheet = 1)


gulf = gulf %>%
  na.omit()



# gulf.tb = gulf %>%
#   mutate(longitude.start = char2dms(from = gulf$`Latitude start`, 
#                                     chd = "°", chm = "'", chs = " ") %>% as.numeric(),
#          longitude.stop = char2dms(from = gulf$`Latitude stop`, 
#                                    chd = "°", chm = "'", chs = " ") %>% as.numeric(),
#          latitude.start = char2dms(from = gulf$`Longitude start`, 
#                                    chd = "°", chm = "'", chs = " ") %>% as.numeric(),
#          latitude.stop = char2dms(from = gulf$`Longitude stop`, 
#                                   chd = "°", chm = "'", chs = " ") %>% as.numeric())

gulf.tb = gulf %>%
  mutate(latitude.start = parzer::parse_lat(`Longitude start`),
         longitude.start = parzer::parse_lon(`Latitude start`),
         latitude.stop = parzer::parse_lat(`Longitude stop`),
         longitude.stop = parzer::parse_lon(`Latitude stop`))
```


```{r tab1}

# gulf.tb %>% select(1:7) %>%
#   mutate(Date = lubridate::as_date(Date))%>%
#   mutate_if(is.numeric, round, digits = 3)%>%DT::datatable(caption = "The 2014 sampled stations with corresponding geographical positions presented in decimal minutes and seconds")

gulf.tb %>% select(1:7) %>%
  kableExtra::kable(format = "html", 
                    col.names = c("Trawl", "Station", "Date", "Latitude", "Longitude", "Latitude","Longitude"),
                    caption = "The 2014 sampled stations with corresponding geographical positions presented in decimal degree") %>%
  kableExtra::column_spec(column = 1:7,width = "3cm") %>%
  kableExtra::add_header_above(c("","","", "Coordinates Start"=2, "Coordinates Stop"=2)) %>%
  kableExtra::add_header_above(c("","","", "Geographical coordinates (DMS)" = 4))
```

## Coordinate Error
It's a common mistake when data is entered manual to have incorrectly labels or missing sign for longitude and latituve value [@parzer]. The data was first explored to identify the correct columns and label their corresponding variables [@tidyverse]. Once the data was assigned to the correct names, was converted from degree--minutes--seconds to decimal degress---a format readable to the computer [Table \@ref(tab:tab2)]



```{r tab2 }
# gulf.tb %>% select(1:3,8:11) %>%
#   mutate(Date = lubridate::as_date(Date))%>%
#   mutate_if(is.numeric, round, digits = 5)%>%DT::datatable(caption = "The 2014 sampled stations with corresponding geographical positions presented in decimal degree") 

gulf.tb %>% select(1:3,8:11) %>%
  kableExtra::kable(format = "html", 
                    col.names = c("Trawl", "Station", "Date", "Latitude", "Longitude", "Latitude","Longitude"),
                    caption = "The 2014 sampled stations with corresponding geographical positions presented in decimal degree") %>%
  kableExtra::column_spec(column = 1:7,width = "2cm") %>%
  kableExtra::add_header_above(c("","","", "Coordinates Start"=2, "Coordinates Stop"=2)) %>%
  kableExtra::add_header_above(c("","","", "Geographical coordinates (DD)" = 4))
```

## Mapped 2014 Locations
Oncce the coordinates were cleaned and arranged in the format that machine readable [@sf], they were mapped [@leaflet] as shown in figure \@ref(fig:fig98). There are few records just before the Kissenda Island and most of the locations are clustered in Butimba Bay. Across the channel is black without a single location. Some locations should in the lake but are on the land instead. *Any idea why that may have happened?* I suspected that the converstion of the data from degree--minutes--seconds to decimal--degree could contributed to this error! However, that wasn't an issue. We will have to check on the original data if it does exist somewhere!

```{r fig98, fig.cap="The map showing the positions of the sampled locations across the Mwanza Gulf recorded in 2014. To view the name of the marks on the map just click the center of the mark"}
gulf.tb %>%
  leaflet() %>%
  addTiles() %>%
  # addMarkers(lng = ~longitude.start, lat = ~latitude.start, popup = ~`Trawling number`)%>%
  addMarkers(lng = ~longitude.stop, lat = ~latitude.stop, popup = ~Station)


```

## Sampled Stations 2006
The second source of the data was collected in 2006 and crossed the same path that the 2014 locations data were marked. Table \@ref(tab:tab3) show the geographical positions along the transect both in original format (degree--minutes) and after converted to decimal degree format. 

```{r }
gulf2 = readxl::read_excel("Coordinates Mwanza Gulf_2006, 2008, 2014-2.xlsx", sheet = 2) 


# gulf2.tb = gulf2 %>%
#   mutate(lon.dd = char2dms(from = lon, chd = "°", chm = "'") %>% as.numeric(),
#          lat.dd = char2dms(from = lat, chd = "°", chm = "'") %>% as.numeric())

gulf2.tb = gulf2 %>%
  mutate(lon.dd = parzer::parse_lon(lon),
         lat.dd = parzer::parse_lat(lat))
```

```{r tab3}
# gulf2.tb %>%
#   mutate_if(is.numeric, round, digits = 5)%>%DT::datatable(rownames = FALSE, colnames = c("Station", "Latitude_DMS", "Longitude_DMS", "Longitude_DD", "Latitude_DD"), caption = "The 2006 sampled stations with corresponding geographical positions both in Decimal minutes and decimal degree")

gulf2.tb %>%
  kableExtra::kable(format = "html", col.names = c("Station", "Latitude", "Longitude", "Longitude", "Latitude"), caption = "The 2006 sampled stations with corresponding geographical positions both in Decimal minutes and decimal degree") %>%
  kableExtra::column_spec(column = 1:5,width = "3cm") %>%
  kableExtra::add_header_above(c("", "Degree Minutes"=2, "Decimal Degree"=2)) %>%
  kableExtra::add_header_above(c("", "Geographical coordinates" = 4))
```

```{r}
gulf2.sf = gulf2.tb %>% st_as_sf(coords = c("lon.dd", "lat.dd")) %>%st_set_crs(4326) 
```

## Mapped 2006 Locations
Unlike the 2014 dataset, the 2006 dataset has `r nrow(gulf2.tb)` points compared to `r nrow(gulf.tb)` of 2014. Despite the few points, the 2006 geographical positions (Table \@ref(tab:tab3)) align along the sampled area with an inverval of about `r gulf2.sf%>% st_distance() %>% mean() %>% as.numeric()` from one point to the next (Figure \@ref(fig:fig99)) 


```{r fig99, fig.cap="The map showing the positions of the sampled locations across the Mwanza Gulf recorded in 2006. To view the name of the marks on the map just click the center of the mark"}
gulf2.tb %>%
  leaflet() %>%
  addTiles() %>%
  addMarkers(lng = ~lon.dd, lat = ~lat.dd, popup = ~Station)
```


```{r, eval=FALSE}
require(ggspatial)
require(ggrepel)

gulf2.tb %>% 
  st_as_sf(coords = c("lon.dd", "lat.dd")) %>% 
  st_set_crs(4326) %>%
  ggplot(aes(x = lon, y = lat)) +
  ggspatial::annotation_map_tile(type = "osm") +
  geom_sf() +
  ggspatial::stat_spatial_identity(aes(label = Station), geom = "text_repel")
  
```

# References{}